---
title: "06_Network_analysis_with_NetCoMi"
output: html_document
date: "2025-12-19"
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "~/ABROGATE")
```

### Introduction
This analysis constructs microbial association networks using the NetCoMi package to compare microbiome compositions between healthy (non-IgE) individuals and those with food allergies. We apply SPIEC-EASI (SParse InversE Covariance Estimation for Ecological Association Inference) with the graphical lasso method to infer microbial co-occurrence and co-exclusion patterns.

The analysis focuses on three comparisons:
- Non-IgE vs. all IgE-mediated food allergy samples
- Non-IgE vs. Cluster B (specific allergic subgroup)
- Non-IgE vs. Cluster E (specific allergic subgroup)

### Load Required Libraries
```{r}
library(magrittr) # needs to be run every time you start R and want to use %>%
library(dplyr) # alternatively, this also loads %>%
library(tidyr)
library(NetCoMi)
library(limma)
```

### Load Count Data and Construct Networks

We load taxonomic count data aggregated at the family level for each sample group. The data represents microbial family abundances across healthy and allergic individuals, with separate files for the full allergic cohort and specific patient clusters

```{r}
## 1. Load your specific data
data_files <- list(
  healthy = 'data/counts_healthy_family.csv',
  allergic = 'data/counts_allergic_family.csv',
  B = 'data/counts_B_family.csv',
  E = 'data/counts_E_family.csv'
)

# Load all datasets in one loop
raw_counts <- lapply(data_files, function(file) {
  t(read.csv(file, row.names = 1, check.names = FALSE))
})

# Verification
cat("Bata dimensions:\n")
for(name in names(raw_counts)) {
  cat(sprintf("%s group: %d samples x %d taxa\n",
              name, nrow(raw_counts[[name]]), ncol(raw_counts[[name]])))
}

## 2. Network Construction
# Befine common parameters once
common_params <- list(
  measure = "spieceasi",
  measurePar = list(
    method = "glasso",
    nlambda = 30,
    lambda.min.ratio = 0.01,
    pulsar.params = list(rep.num = 20)
  ),
  filtSamp = "none",
  filtTax = "highestVar",
  filtTaxPar = list(highestVar = 50),
  normMethod = "none",
  sparsMethod = "none",
  verbose = 3,
  seed = 123456
)

# Function to construct network with common parameters
construct_net <- function(data1, data2, params) {
  do.call(netConstruct, c(list(data = data1, data2 = data2), params))
}

# Construct all networks
networks <- list(
  any = construct_net(raw_counts$healthy, raw_counts$allergic, common_params),
  B = construct_net(raw_counts$healthy, raw_counts$B, common_params),
  E = construct_net(raw_counts$healthy, raw_counts$E, common_params)
)

# Create copies of all networks
networks_copy <- networks

# Assign to individual variables
net_any <- networks$any
net_B <- networks$B
net_E <- networks$E

# Individual copies (if you need to modify them separately)
net_any_copy <- networks_copy$any
net_B_copy <- networks_copy$B
net_E_copy <- networks_copy$E
```

Since cluster B and E have different microbial compositions than the full allergic cohort, the variance profiles differ, resulting in different taxa sets. Even though SPIEC-EASI estimates the healthy network independently in each case, it operates on different taxa subsets, producing different edge structures.

### Network Visualization Setup

We assign colors to bacterial phyla using a colorblind-friendly palette to ensure accessibility in visualizations [cite:5]. Each microbial family is colored according to its phylum membership for consistent interpretation across all network plots.

```{r}
## Load phylum labels and create color mapping
p_labels <- read.csv('data/phylum_labels_only_taxa.csv',
                     row.names = 1, check.names = FALSE)

cat("Phylum label structure:\n")
print(head(p_labels))
cat("\nUnique phyla:\n")
print(table(p_labels$phylum))

# Colorblind-friendly palette with 13 colors
phylcolors <- c("#88CCEE", "#CC6677", "#BBCC77", "#117733", "#332288", 
                "#AA4499", "#44AA99", "#999933", "#882255", "#661100", 
                "#6699CC", "#888888", "#E69F00")

# Get all unique phyla from the complete dataset
unique_phyla <- sort(unique(p_labels$phylum))

# Check we have enough colors
if(length(unique_phyla) > length(phylcolors)) {
  stop("Not enough colors for all phyla!")
}

# Trim and name colors
phylcolors <- phylcolors[1:length(unique_phyla)]
names(phylcolors) <- unique_phyla

cat("\nPhylum to color mapping:\n")
print(phylcolors)

## Create a function to map taxa to colors
get_node_colors <- function(taxa_names, p_labels, phylcolors) {
  phyla <- p_labels[taxa_names, "phylum"]
  colors <- phylcolors[phyla]
  names(colors) <- taxa_names
  return(colors)
}

## Create output directories
dir.create("plots/png", showWarnings = FALSE, recursive = TRUE)
dir.create("plots/svg", showWarnings = FALSE, recursive = TRUE)

# Network configurations
network_configs <- list(
  any = list(
    net = net_any,
    filename_png = "plots/png/partial_corr_allergic_vs_healthy.png",
    filename_svg = "plots/svg/partial_corr_allergic_vs_healthy.svg",
    groups = c("non-IgE", "IgE")
  ),
  B = list(
    net = net_B,
    filename_png = "plots/png/partial_corr_B_vs_healthy.png",
    filename_svg = "plots/svg/partial_corr_B_vs_healthy.svg",
    groups = c("non-IgE", "Cluster B")
  ),
  E = list(
    net = net_E,
    filename_png = "plots/png/partial_corr_E_vs_healthy.png",
    filename_svg = "plots/svg/partial_corr_E_vs_healthy.svg",
    groups = c("non-IgE", "Cluster E")
  )
)

## STEP 1: Find common taxa
cat("\nFinding common taxa across all healthy networks...\n")

common_taxa_healthy <- rownames(network_configs$any$net$adjaMat1)
for(name in names(network_configs)) {
  taxa_current <- rownames(network_configs[[name]]$net$adjaMat1)
  common_taxa_healthy <- intersect(common_taxa_healthy, taxa_current)
}

cat("Common taxa in all healthy networks:", length(common_taxa_healthy), "\n")

## STEP 2: Create reference layout
cat("\nCreating reference layout from first network (common taxa only)...\n")

net_any_common <- net_any
net_any_common$adjaMat1 <- net_any$adjaMat1[common_taxa_healthy, common_taxa_healthy]
net_any_common$adjaMat2 <- net_any$adjaMat2[common_taxa_healthy, common_taxa_healthy]
net_any_common$assoMat1 <- net_any$assoMat1[common_taxa_healthy, common_taxa_healthy]
net_any_common$assoMat2 <- net_any$assoMat2[common_taxa_healthy, common_taxa_healthy]

props_any <- netAnalyze(net_any_common)

# Get all nodes and assign colors
all_nodes_any <- union(rownames(net_any_common$adjaMat1), rownames(net_any_common$adjaMat2))
node_colors_any <- get_node_colors(all_nodes_any, p_labels, phylcolors)

# Create phylum vector for legend
phylum_vector_common <- p_labels[all_nodes_any, "phylum"]
names(phylum_vector_common) <- all_nodes_any

p_any <- plot(props_any, 
              groupNames = network_configs$any$groups,
              sameLayout = TRUE,
              rmSingles = "inboth",
              nodeColor = "colorVec",
              colorVec = node_colors_any,
              featVecCol = phylum_vector_common,
              legendArgs = list(title = "Phylum"),
              repulsion = 0.9,
              labelScale = FALSE,
              cexLabels = 1.0,
              labelCol = "black")

reference_layout <- p_any$layout$layout1
reference_taxa <- rownames(reference_layout)

cat("Reference layout created with", nrow(reference_layout), "nodes\n")

## STEP 3: Save first network plots
results <- list()

cat("\nSaving plots for network: any (IgE vs non-IgE)\n")

# PNG - 4000 x 2500 pixels at 300 BPI = 13.33 x 8.33 inches
png(network_configs$any$filename_png, width = 4000, height = 2500, res = 300)
plot(props_any, 
     groupNames = network_configs$any$groups,
     sameLayout = TRUE,
     rmSingles = "inboth",
     nodeColor = "colorVec",
     colorVec = node_colors_any,
     featVecCol = phylum_vector_common,
     legendArgs = list(title = "Phylum"),
     repulsion = 0.9,
     labelScale = FALSE,
     cexLabels = 1.0,
     labelCol = "black")
dev.off()
cat(sprintf("  PNG saved: %s\n", network_configs$any$filename_png))

# SVG - same dimensions as PNG: 13.33 x 8.33 inches
svg(network_configs$any$filename_svg, width = 13.33, height = 8.33)
plot(props_any, 
     groupNames = network_configs$any$groups,
     sameLayout = TRUE,
     rmSingles = "inboth",
     nodeColor = "colorVec",
     colorVec = node_colors_any,
     featVecCol = phylum_vector_common,
     legendArgs = list(title = "Phylum"),
     repulsion = 0.9,
     labelScale = FALSE,
     cexLabels = 1.0,
     labelCol = "black")
dev.off()
cat(sprintf("  SVG saved: %s\n", network_configs$any$filename_svg))

# Create differential network for 'any'
taxa_any <- rownames(p_any$layout$layout1)
net_any_filtered <- net_any_common
net_any_filtered$assoMat1 <- net_any_common$assoMat1[taxa_any, taxa_any]
net_any_filtered$assoMat2 <- net_any_common$assoMat2[taxa_any, taxa_any]
net_any_filtered$assoEst1 <- net_any_common$assoEst1[taxa_any, taxa_any]
net_any_filtered$assoEst2 <- net_any_common$assoEst2[taxa_any, taxa_any]

diff_any <- diffnet(net_any_filtered, diffMethod = "threshold", 
                    diffThresh = 0.0, adjust = "none")

results$any <- list(props = props_any, plot = p_any, diff = diff_any, 
                    net_filtered = net_any_filtered, 
                    node_colors = node_colors_any,
                    phylum_vector = phylum_vector_common)

# Process B and E networks
for(name in c("B", "E")) {
  config <- network_configs[[name]]
  cat(sprintf("\nProcessing network: %s (using shared non-IgE layout)\n", name))
  
  taxa_in_network <- intersect(reference_taxa, rownames(config$net$adjaMat1))
  
  if(length(taxa_in_network) < length(reference_taxa)) {
    cat(sprintf("  Warning: Only %d/%d reference taxa found\n", 
                length(taxa_in_network), length(reference_taxa)))
  }
  
  net_filtered <- config$net
  net_filtered$adjaMat1 <- config$net$adjaMat1[taxa_in_network, taxa_in_network]
  net_filtered$adjaMat2 <- config$net$adjaMat2[taxa_in_network, taxa_in_network]
  net_filtered$assoMat1 <- config$net$assoMat1[taxa_in_network, taxa_in_network]
  net_filtered$assoMat2 <- config$net$assoMat2[taxa_in_network, taxa_in_network]
  
  props <- netAnalyze(net_filtered)
  
  degree1 <- rowSums(net_filtered$adjaMat1 != 0)
  degree2 <- rowSums(net_filtered$adjaMat2 != 0)
  has_edges <- (degree1 > 0) | (degree2 > 0)
  taxa_with_edges <- taxa_in_network[has_edges]
  
  cat(sprintf("  Removed %d singletons\n", sum(!has_edges)))
  
  layout_subset <- reference_layout[taxa_with_edges, , drop = FALSE]
  
  net_filtered$adjaMat1 <- net_filtered$adjaMat1[taxa_with_edges, taxa_with_edges]
  net_filtered$adjaMat2 <- net_filtered$adjaMat2[taxa_with_edges, taxa_with_edges]
  net_filtered$assoMat1 <- net_filtered$assoMat1[taxa_with_edges, taxa_with_edges]
  net_filtered$assoMat2 <- net_filtered$assoMat2[taxa_with_edges, taxa_with_edges]
  
  props <- netAnalyze(net_filtered)
  
  # Get node colors for this network
  all_nodes <- union(rownames(net_filtered$adjaMat1), rownames(net_filtered$adjaMat2))
  node_colors <- get_node_colors(all_nodes, p_labels, phylcolors)
  
  # Create phylum vector for legend
  phylum_vector <- p_labels[all_nodes, "phylum"]
  names(phylum_vector) <- all_nodes
  
  # PNG
  png(config$filename_png, width = 4000, height = 2500, res = 300)
  p <- plot(props, 
            groupNames = config$groups,
            layout = layout_subset,
            sameLayout = TRUE,
            rmSingles = FALSE,
            nodeColor = "colorVec",
            colorVec = node_colors,
            featVecCol = phylum_vector,
            legendArgs = list(title = "Phylum"),
            repulsion = 0.9,
            labelScale = FALSE,
            cexLabels = 1.0,
            labelCol = "black")
  dev.off()
  cat(sprintf("  PNG saved: %s\n", config$filename_png))
  
  # SVG - same dimensions as PNG
  svg(config$filename_svg, width = 13.33, height = 8.33)
  p <- plot(props, 
            groupNames = config$groups,
            layout = layout_subset,
            sameLayout = TRUE,
            rmSingles = FALSE,
            nodeColor = "colorVec",
            colorVec = node_colors,
            featVecCol = phylum_vector,
            legendArgs = list(title = "Phylum"),
            repulsion = 0.9,
            labelScale = FALSE,
            cexLabels = 1.0,
            labelCol = "black")
  dev.off()
  cat(sprintf("  SVG saved: %s\n", config$filename_svg))
  
  # Bifferential network
  taxa_plot <- rownames(p$layout$layout1)
  net_diff <- net_filtered
  net_diff$assoMat1 <- net_filtered$assoMat1[taxa_plot, taxa_plot]
  net_diff$assoMat2 <- net_filtered$assoMat2[taxa_plot, taxa_plot]
  net_diff$assoEst1 <- net_filtered$assoEst1[taxa_plot, taxa_plot]
  net_diff$assoEst2 <- net_filtered$assoEst2[taxa_plot, taxa_plot]
  
  diff_net <- diffnet(net_diff, diffMethod = "threshold", 
                      diffThresh = 0.0, adjust = "none")
  
  results[[name]] <- list(props = props, plot = p, diff = diff_net, 
                          net_filtered = net_diff, 
                          node_colors = node_colors,
                          phylum_vector = phylum_vector)
}

# Access results
props_any <- results$any$props
p_any <- results$any$plot
diff_any <- results$any$diff

props_B <- results$B$props
p_B <- results$B$plot
diff_B <- results$B$diff

props_E <- results$E$props
p_E <- results$E$plot
diff_E <- results$E$diff

cat("\nAll networks processed successfully!\n")
cat("PNG plots saved to plots/png/\n")
cat("SVG plots saved to plots/svg/\n")
```

### Differential Network Analysis

Differential network analysis identifies changes in microbial associations between healthy and allergic groups [cite:6]. Each edge in the differential network represents a change in association strength or direction between the two conditions. This reveals which microbial relationships are gained, lost, strengthened, weakened, or reversed in the context of food allergy.

#### Generate Differential Network Plots
We use a consistent spatial layout across all comparisons to facilitate visual comparison. Nodes (bacterial families) maintain the same positions, while edge colors indicate the type and direction of association changes [cite:7].

```{r}
## 4. Plot Bifferential Networks with Shared Layout

# Befine differential plot parameters
diff_plot_params <- list(
  rmSingles = FALSE,
  mar = c(2, 2, 4, 10),
  edgeCol = c("#CC79A7",  # + | + (both positive, stronger in group 2) - reddish purple
            "#009E73",  # + | 0 (positive to zero) - teal/green
            "#0072B2",  # + | - (positive to negative) - blue
            "#E69F00",  # - | + (negative to positive) - orange
            "#999999",  # - | 0 (negative to zero) - gray
            "#56B4E9",  # - | - (both negative) - light blue (was white)
            "#F0E442",  # 0 | + (zero to positive) - yellow
            "white",  # 0 | 0 (no edge in either)
            "#B55E00"),  # 0 | - (zero to negative) - red-orange

  edgeWidth = 2.0,
  labelScale = FALSE,
  cexLabels = 1.0
)

# Function to plot differential network with proper layout matching
plot_diffnet_safe <- function(diff_obj, base_plot,
                              filename_png, filename_svg, title, plot_params) {
  
  # Use the exact layout from the corresponding base network plot
  layout_to_use <- base_plot$layout$layout1
  
  cat(sprintf("  Base network layout: %d nodes\n", nrow(layout_to_use)))
  
  # PNG plot
  png(filename_png, width = 4000, height = 2500, res = 300)
  plot(diff_obj, 
       layout = layout_to_use, 
       main = title,
       rmSingles = FALSE,
       mar = plot_params$mar,
       edgeCol = plot_params$edgeCol,
       edgeWidth = plot_params$edgeWidth,
       labelScale = plot_params$labelScale,
       cexLabels = plot_params$cexLabels,
       labelCol = "black")
  dev.off()
  
  # SVG plot
  svg(filename_svg, width = 13.33, height = 8.33)
  plot(diff_obj, 
       layout = layout_to_use, 
       main = title,
       rmSingles = FALSE,
       mar = plot_params$mar,
       edgeCol = plot_params$edgeCol,
       edgeWidth = plot_params$edgeWidth,
       labelScale = plot_params$labelScale,
       cexLabels = plot_params$cexLabels,
       labelCol = "black")
  dev.off()
  
  cat(sprintf("  PNG saved: %s\n", filename_png))
  cat(sprintf("  SVG saved: %s\n", filename_svg))
}

# Plot all differential networks using their corresponding base network layouts
cat("\nPlotting differential networks with matched layouts...\n")

# Plot differential network for 'any'
cat("\nProcessing differential network: any (non-IgE vs IgE)\n")
plot_diffnet_safe(
  diff_obj = results$any$diff,
  base_plot = results$any$plot,
  filename_png = "plots/png/differential_allergic_vs_healthy.png",
  filename_svg = "plots/svg/differential_allergic_vs_healthy.svg",
  title = "Bifferential Network: non-IgE vs IgE",
  plot_params = diff_plot_params
)

# Plot differential network for 'B'
cat("\nProcessing differential network: B (non-IgE vs Cluster B)\n")
plot_diffnet_safe(
  diff_obj = results$B$diff,
  base_plot = results$B$plot,
  filename_png = "plots/png/differential_B_vs_healthy.png",
  filename_svg = "plots/svg/differential_B_vs_healthy.svg",
  title = "Bifferential Network: non-IgE vs Cluster B",
  plot_params = diff_plot_params
)

# Plot differential network for 'E'
cat("\nProcessing differential network: E (non-IgE vs Cluster E)\n")
plot_diffnet_safe(
  diff_obj = results$E$diff,
  base_plot = results$E$plot,
  filename_png = "plots/png/differential_E_vs_healthy.png",
  filename_svg = "plots/svg/differential_E_vs_healthy.svg",
  title = "Bifferential Network: non-IgE vs Cluster E",
  plot_params = diff_plot_params
)

cat("\nAll differential network plots completed!\n")
cat("PNG plots saved to plots/png/\n")
cat("SVG plots saved to plots/svg/\n")
cat("\nNOTE: Only edges showing significant differences are displayed (colored).\n")
cat("Grey edges (non-significant differences) are hidden.\n")

```

### Create Edge Change and Phylum Legends

The edge legend explains how microbial associations change between healthy and allergic states, while the phylum legend provides taxonomic context for node colors.

```{r}
# Befine edge colors (matching your differential network settings)
edgeCol = c("#CC79A7",  # + | + (both positive, stronger in group 2) - reddish purple
            "#009E73",  # + | 0 (positive to zero) - teal/green
            "#0072B2",  # + | - (positive to negative) - blue
            "#E69F00",  # - | + (negative to positive) - orange
            "#999999",  # - | 0 (negative to zero) - gray
            "#56B4E9",  # - | - (both negative) - light blue (was white)
            "#F0E442",  # 0 | + (zero to positive) - yellow
            "white",  # 0 | 0 (no edge in either)
            "#B55E00")  # 0 | - (zero to negative) - red-orange

# Filter to show only visible (non-white) edges
visible_edges <- edgeCol != "white"
edgeCol_visible <- edgeCol[visible_edges]
edge_labels <- c(
              "+ → + (stronger positive in group 2)",
              "+ → 0 (association lost)",
              "+ → − (association reversed)",
              "− → + (association reversed)",
              "− → 0 (negative association lost)",
              "− → − (stronger negative in group 2)",
              "0 → + (new positive association)",
              "0 → 0 (no edge in either group)",
              "0 → − (new negative association)")

# Edge type legend - PNG
png("plots/png/differential_edge_legend.png", 
    width = 1500, height = 2000, res = 300)

par(mar = c(1, 1, 2, 1))
plot.new()

legend("center",
       title = "Edge Changes (Non-IgE → IgE)",
       legend = edge_labels,
       col = edgeCol_visible,
       lty = 1,
       lwd = 5,
       cex = 1.8,
       bty = "n",
       title.cex = 2.0)

dev.off()

# Edge type legend - SVG
svg("plots/svg/differential_edge_legend.svg", 
    width = 5, height = 6.67)

par(mar = c(1, 1, 2, 1))
plot.new()

legend("center",
       title = "Edge Changes (Non-IgE → IgE)",
       legend = edge_labels,
       col = edgeCol_visible,
       lty = 1,
       lwd = 5,
       cex = 1.8,
       bty = "n",
       title.cex = 2.0)

dev.off()

cat("Edge legend saved to PNG and SVG\n")

# Phylum legend - PNG
png("plots/png/phylum_legend.png", width = 1200, height = 2000, res = 300)

par(mar = c(1, 1, 2, 1))
plot.new()

# Get phylum vector from one of the networks (use 'any' network)
phylum_vector <- results$any$phylum_vector
phyla_present <- sort(unique(phylum_vector))
phyla_clean <- gsub("p__|;", "", phyla_present)
colors_present <- phylcolors[phyla_present]

legend("center",
       title = "Bacterial Phyla",
       legend = phyla_clean,
       fill = colors_present,
       cex = 2.0,
       bty = "n",
       title.cex = 2.5)

dev.off()

# Phylum legend - SVG
svg("plots/svg/phylum_legend.svg", width = 4, height = 6.67)

par(mar = c(1, 1, 2, 1))
plot.new()

legend("center",
       title = "Bacterial Phyla",
       legend = phyla_clean,
       fill = colors_present,
       cex = 2.0,
       bty = "n",
       title.cex = 2.5)

dev.off()

cat("Phylum legend saved to PNG and SVG\n")
cat("\nLegend files saved:\n")
cat("  PNG: plots/png/differential_edge_legend.png\n")
cat("  SVG: plots/svg/differential_edge_legend.svg\n")
cat("  PNG: plots/png/phylum_legend.png\n")
cat("  SVG: plots/svg/phylum_legend.svg\n")
```

### Summary

This analysis generated microbial association networks comparing healthy individuals to allergic patients and specific allergic subgroups. The networks reveal structural differences in microbial co-occurrence patterns associated with food allergy phenotypes. 

Key outputs include:
- Three paired network comparisons (non-IgE vs. IgE, Cluster B, and Cluster E)
- Differential network plots highlighting association changes
- Standalone legend files for interpretation
